Longitudinal Manifold Estimation

### 2023-02-14 paper/main.tex
begin introduction rewrite

@@

Because the algorithm proposed in this article relies on the same definitions and notation as those introduced in Meng and Eloyan (2021), we provide it here. To generalize the principal curve framework proposed in Hastie and Stuetzle (1989) to intrinsic dimensions $d > 1$, Meng and Eloyan introduce the function space 
\[%
  C_{\infty}(\mathbb{R}^d \to \mathbb{R}^D) = \left\{f \in C()\right\}
.\]%

@@

\begin{algorithm}
\caption{Longitudinal Principal Manifold Estimation}\label{alg:two}
\KwData{Data points $\{y_{i, t}\}_{i, t}^{I, T}$, positive integer $d$, positive integer $N_0 < I - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\{\lambda_k\}_{k=1}^K$, $\{\gamma_l\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $g^*: \mathbb{R}^{d+1} \to \mathbb{R}^{D+1}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, ... T$}{
    Apply PME algorithm with input $(\{y_{i, t}\}_{i=1}^I, N_0, \alpha, \epsilon, \epsilon^*, \{\lambda_k\}_{k=1}^K)$ and obtain $\{f_t^*\}_{t=1}^T$ and $\{\lambda_t^*\}_{t=1}^T$\;
    Use $\{f_t^*\}_{t=1}^T$ to map a grid of $d$-dimensional parameters $\{r_j\}_{j=1}^N$ to $\{x_{j, t}\}_{j=1}^N$ in $\mathbb{R}^D$\;
}
Apply ISOMAP to parameterize the newly generated collection $\{x_{j, t}\}_{j, t}^{N, T}$ by the $d$-dimensional parameters $\{r_j\}_{j=1}^N$\;
Use parameters $\{r_j\}_{j=1}^N$ and time values $\{t_j\}_{j=1}^N$ to create new
$\pi_{g_{(1)}}(x_{j}) \gets r_j$ for $j=1, ..., N$\;
\For{$l = 1, 2, ..., L$}{
    $\gamma \gets \gamma_l$.\;
    Obtain $g_{(1)}$ by solving 
    $\left(\begin{array}{ccc} 
    2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
    2\mathbf{T}^T\mathbf{W}\mathbf{E} & 2\mathbf{T}^T\mathbf{W}\mathbf{T} & \mathbf{0} \\
    \mathbf{T}^T & \mathbf{0} & \mathbf{0} \end{array}\right)\left(\begin{array}{c} 
    s_l \\
    \alpha_l \\
    m_l 
    \end{array}\right) = \left(\begin{array}{c}
    2\mathbf{E}\mathbf{W}\mathbf{x} \\
    2\mathbf{T}^T\mathbf{W}\mathbf{x} \\
    \mathbf{0} 
    \end{array}\right)$\;
    $\mathcal{E} \gets 2\epsilon^*$\;
    $n \gets 1$\;
    $\mathcal{D}_{\hat{Q}, N}(g_{(1)}) \gets \sum_{j=1}^N\|x_j - g_{(1)}\|_{\mathbb{R}^{D}}^2$\;
    \While{$\mathcal{E} \geq \epsilon^*$, $n < itr$}{
        Compute $g_{(n + 1)}$ by solving $\left(\begin{array}{ccc} 
        2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
        2\mathbf{T}^T\mathbf{W}\mathbf{E} & 2\mathbf{T}^T\mathbf{W}\mathbf{T} & \mathbf{0} \\
        \mathbf{T}^T & \mathbf{0} & \mathbf{0} \end{array}\right)\left(\begin{array}{c} 
        s_l \\
        \alpha_l \\
        m_l 
        \end{array}\right) = \left(\begin{array}{c}
        2\mathbf{E}\mathbf{W}\mathbf{x} \\
        2\mathbf{T}^T\mathbf{W}\mathbf{x} \\
        \mathbf{0} 
        \end{array}\right)$\;
        $\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) \gets \sum_{j=1}^N\|x_j - g_{(n + 1)}(\pi_{g_{(n + 1)}}(x_j))\|_{\mathbb{R}^D}^2$\;
        $\mathcal{E} \gets |\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) - \mathcal{D}_{\hat{Q}_N}(g_{(n)})|/\mathcal{D}_{\hat{Q}_N}(g_{(n)})$\;
        $n \gets n + 1$\;
    }
    $\hat{g}_l \gets g_{(n)}$\;
}
$\kappa \gets \text{max}\{\|\pi_{\hat{g}_{l^*}}(x_j)\|_{\mathbb{R}^d}: j = 1, 2, \dots, I\}$, where $l^* = \arg \min_l\{\frac{1}{N}\sum_{j=1}^N\|x_j - g_{(l)}(\pi_{\hat{g}_{(l)}}(x_j)\|_{\mathbb{R}^D}^2: l = 1, 2, \dots, L\}$\;
$g^*(r) = g^{l^*}(\kappa r)$, where the analytic formula of $g^*$ is from $g_{(n + 1), l}(r) = \sum_{j=1}^N \eta_{4-d}(r - \pi_{g_{(n)}}(x_j)) + \sum_{k=1}^{d + 1} \alpha_{k, l} \times p_k(r)$, and $\gamma^* \gets \gamma_{l^*}$
\end{algorithm}

	\begin{equation}\label{PMSEF}
	\mathcal{K}_{\lambda,\mathbb{P}}(f,g)=\mathbb{E}\left\Vert X_t - f\left(\pi_g(X_t)\right)\right\Vert^2_{\mathbb{R}^D} + \lambda \sum_{t = 1}^T \left\Vert\nabla^{\otimes 2} f(t)\right\Vert_{L^2(\mathbb{R}^d)}^2 + \gamma \sum_{X} \left\Vert\nabla^{\otimes 2} f(X_t)\right\Vert_{L^2(\mathbb{R})}^2, 
	\end{equation}
	\begin{equation}\nonumber
	\mathcal{K}_{\lambda,\mathbb{P}}(f)=\mathcal{K}_{\lambda,\mathbb{P}}(f,f),
	\end{equation}
	where $\Vert\nabla^{\otimes 2}f \Vert_{L^2(\mathbb{R}^d)}^2 = \sum_{l=1}^D\int_{\mathbb{R}^d} \sum_{i,j=1}^d\vert\frac{\partial^2 f_l}{\partial t_i \partial t_j}(t)\vert^2dt$. A manifold $M_{f^*}^d$ determined by $f^*$ is called a principal manifold for $X$ (or $\mathbb{P}$) with the tuning parameter $\lambda$ if 
	\begin{align}\label{def: PM}
	f^*=\arg\min_{f\in\mathscr{F}(\mathbb{P})}\mathcal{K}_{\lambda,\mathbb{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbb{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbb{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
	\end{align}

>>>>

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (Knopman et al (2021), Poewe et al (2017)). Frequently, interest centers around changes in one or more substructures. For example, it is common to observe atrophy in the hippocampus of those with Alzheimer's disease or cognitive impairment from other causes. Image segmentation allows for the extraction and analysis of these subcortical structures.

Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach, and is the standard. However, this approach is highly time and resource intensive. In studies analyzing a large number of images, this may be cost prohibitive. Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability. Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns. While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

To demonstrate the potential extent of the inaccuracies introduced by segmentation, using FSL-FIRST to segment the hippocampus in images of one healthy individual over the course of four year in the ADNI dataset, the volume of the left and right hippocampi decreased by 3.6\% and 5.4\%, respectively, while each hippocampus was estimated to have increased in volume in subsequent images twice. In a larger comparison of the accuracy of FSL-FIRST and FreeSurfer, Mulder et al. (2014) found that 6.9\% of segmentations by FIRST failed visual inspection, as did 7.5\% of segmentations for FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with large numbers of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation. Mulder et al. (2014) also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches.

Ultimately, high levels of variability are present at the subcortical level regardless of the segmentation approach used, but may be particularly influential when using automated segmentation methods. To address these elevated noise levels, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time. This approach, which extends the principal manifold estimation algorithm introduced by Meng and Eloyan (2021) for a longitudinal setting, includes a regularization parameter to limit the effects of between time-point variability.

\section{Introduction - Old}

@@

Because the algorithm proposed in this article relies on the same definitions and notation as those introduced in Meng and Eloyan (2021), we provide it here. To generalize the principal curve framework proposed in Hastie and Stuetzle (1989) to intrinsic dimensions $d > 1$, Meng and Eloyan introduce the function space 
\[%
  C_{\infty}(\mathbf{R}^d \to \mathbf{R}^D) = \left\{f \in C()\right\}
.\]%

@@

\begin{algorithm}
\caption{Longitudinal Principal Manifold Estimation}\label{alg:two}
\KwData{Data points $\{y_{i, t}\}_{i, t}^{I, T}$, positive integer $d$, positive integer $N_0 < I - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\{\lambda_k\}_{k=1}^K$, $\{\gamma_l\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $g^*: \mathbf{R}^{d+1} \to \mathbf{R}^{D+1}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, ... T$}{
    Apply PME algorithm with input $(\{y_{i, t}\}_{i=1}^I, N_0, \alpha, \epsilon, \epsilon^*, \{\lambda_k\}_{k=1}^K)$ and obtain $\{f_t^*\}_{t=1}^T$ and $\{\lambda_t^*\}_{t=1}^T$\;
    Use $\{f_t^*\}_{t=1}^T$ to map a grid of $d$-dimensional parameters $\{r_j\}_{j=1}^N$ to $\{x_{j, t}\}_{j=1}^N$ in $\mathbf{R}^D$\;
}
Apply ISOMAP to parameterize the newly generated collection $\{x_{j, t}\}_{j, t}^{N, T}$ by the $d$-dimensional parameters $\{r_j\}_{j=1}^N$\;
Use parameters $\{r_j\}_{j=1}^N$ and time values $\{t_j\}_{j=1}^N$ to create new
$\pi_{g_{(1)}}(x_{j}) \gets r_j$ for $j=1, ..., N$\;
\For{$l = 1, 2, ..., L$}{
    $\gamma \gets \gamma_l$.\;
    Obtain $g_{(1)}$ by solving 
    $\left(\begin{array}{ccc} 
    2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
    2\mathbf{T}^T\mathbf{W}\mathbf{E} & 2\mathbf{T}^T\mathbf{W}\mathbf{T} & \mathbf{0} \\
    \mathbf{T}^T & \mathbf{0} & \mathbf{0} \end{array}\right)\left(\begin{array}{c} 
    s_l \\
    \alpha_l \\
    m_l 
    \end{array}\right) = \left(\begin{array}{c}
    2\mathbf{E}\mathbf{W}\mathbf{x} \\
    2\mathbf{T}^T\mathbf{W}\mathbf{x} \\
    \mathbf{0} 
    \end{array}\right)$\;
    $\mathcal{E} \gets 2\epsilon^*$\;
    $n \gets 1$\;
    $\mathcal{D}_{\hat{Q}, N}(g_{(1)}) \gets \sum_{j=1}^N\|x_j - g_{(1)}\|_{\mathbf{R}^{D}}^2$\;
    \While{$\mathcal{E} \geq \epsilon^*$, $n < itr$}{
        Compute $g_{(n + 1)}$ by solving $\left(\begin{array}{ccc} 
        2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
        2\mathbf{T}^T\mathbf{W}\mathbf{E} & 2\mathbf{T}^T\mathbf{W}\mathbf{T} & \mathbf{0} \\
        \mathbf{T}^T & \mathbf{0} & \mathbf{0} \end{array}\right)\left(\begin{array}{c} 
        s_l \\
        \alpha_l \\
        m_l 
        \end{array}\right) = \left(\begin{array}{c}
        2\mathbf{E}\mathbf{W}\mathbf{x} \\
        2\mathbf{T}^T\mathbf{W}\mathbf{x} \\
        \mathbf{0} 
        \end{array}\right)$\;
        $\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) \gets \sum_{j=1}^N\|x_j - g_{(n + 1)}(\pi_{g_{(n + 1)}}(x_j))\|_{\mathbf{R}^D}^2$\;
        $\mathcal{E} \gets |\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) - \mathcal{D}_{\hat{Q}_N}(g_{(n)})|/\mathcal{D}_{\hat{Q}_N}(g_{(n)})$\;
        $n \gets n + 1$\;
    }
    $\hat{g}_l \gets g_{(n)}$\;
}
$\kappa \gets \text{max}\{\|\pi_{\hat{g}_{l^*}}(x_j)\|_{\mathbf{R}^d}: j = 1, 2, \dots, I\}$, where $l^* = \arg \min_l\{\frac{1}{N}\sum_{j=1}^N\|x_j - g_{(l)}(\pi_{\hat{g}_{(l)}}(x_j)\|_{\mathbf{R}^D}^2: l = 1, 2, \dots, L\}$\;
$g^*(r) = g^{l^*}(\kappa r)$, where the analytic formula of $g^*$ is from $g_{(n + 1), l}(r) = \sum_{j=1}^N \eta_{4-d}(r - \pi_{g_{(n)}}(x_j)) + \sum_{k=1}^{d + 1} \alpha_{k, l} \times p_k(r)$, and $\gamma^* \gets \gamma_{l^*}$
\end{algorithm}

	\begin{equation}\label{PMSEF}
	\mathcal{K}_{\lambda,\mathbf{P}}(f,g)=\mathbf{E}\left\Vert X_t - f\left(\pi_g(X_t)\right)\right\Vert^2_{\mathbf{R}^D} + \lambda \sum_{t = 1}^T \left\Vert\nabla^{\otimes 2} f(t)\right\Vert_{L^2(\mathbf{R}^d)}^2 + \gamma \sum_{X} \left\Vert\nabla^{\otimes 2} f(X_t)\right\Vert_{L^2(\mathbf{R})}^2, 
	\end{equation}
	\begin{equation}\nonumber
	\mathcal{K}_{\lambda,\mathbf{P}}(f)=\mathcal{K}_{\lambda,\mathbf{P}}(f,f),
	\end{equation}
	where $\Vert\nabla^{\otimes 2}f \Vert_{L^2(\mathbf{R}^d)}^2 = \sum_{l=1}^D\int_{\mathbf{R}^d} \sum_{i,j=1}^d\vert\frac{\partial^2 f_l}{\partial t_i \partial t_j}(t)\vert^2dt$. A manifold $M_{f^*}^d$ determined by $f^*$ is called a principal manifold for $X$ (or $\mathbf{P}$) with the tuning parameter $\lambda$ if 
	\begin{align}\label{def: PM}
	f^*=\arg\min_{f\in\mathscr{F}(\mathbf{P})}\mathcal{K}_{\lambda,\mathbf{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbf{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbf{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
	\end{align}
